Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-16 paper.qmd
Moved PBMC3k codes to scripts

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("x", 1:50)

training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
pbmc_scaled_obj <- gen_scaled_data(
  nldr_data = umap_pbmc)
umap_pbmc_scaled <- pbmc_scaled_obj$scaled_nldr
```

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
## Compute hexbin parameters
num_bins_x_pbmc <- 22

algo_obj_pbmc <- fit_highd_model(
  highd_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  bin1 = num_bins_x_pbmc, 
  q = 0.1, 
  benchmark_highdens = 5)

umap_pbmc_scaled <- algo_obj_pbmc$nldr_obj$scaled_nldr
tr_from_to_df_pbmc <- algo_obj_pbmc$trimesh_data
df_bin_centroids_pbmc <- algo_obj_pbmc$model_2d
df_bin_pbmc <- algo_obj_pbmc$model_highd

umap_pbmc_scaled_with_cluster <- umap_pbmc_scaled |>
  mutate(cluster = if_else(UMAP1 <= 0.25, "cluster1", if_else(UMAP1 >= 0.625, "cluster2", "cluster3"))) |>
  mutate(cluster = factor(cluster, levels = c("cluster1", "cluster2", "cluster3")))

trimesh_removed_pbmc <- ggplot() + 
  geom_point(
    data = umap_pbmc_scaled_with_cluster,
    aes(
      x = UMAP1,
      y = UMAP2,
      colour = cluster
    ),
    alpha = 0.2#,
    #size = 0.1#,
    #colour = '#a65628'
  )  +
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("a1", c(0.08, 0.9), cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: prep-pbmc-umap-model-proj

data_pbmc <- training_data_pbmc |> 
  select(-ID) |>
  mutate(type = "data")

# df_bin_centroids_pbmc <- df_bin_centroids_pbmc |>
#   dplyr::filter(std_counts > 0)

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hexID %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hexID),] |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data_pbmc <- bind_rows(data_pbmc, df_b_pbmc)
scaled_pbmc <- scale_data_manual(df_model_data_pbmc, "type") |>
  as_tibble()

scaled_pbmc_data <- scaled_pbmc |>
  filter(type == "data") |>
  select(-type)

scaled_pbmc_data_model <- scaled_pbmc |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-pbmc-umap-author-proj
#| eval: false

data_pbmc_n <- data_pbmc |>
  select(-type) |>
  mutate(type = as.character(umap_pbmc_scaled_with_cluster$cluster))

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc_n)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = tr_from_to_df_pbmc$from,
                         lineTo = tr_from_to_df_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

## First projection
projection <- cbind(
    c(-0.4545,-0.5533,-0.1235,0.3984,-0.2065,-0.3586,-0.0088,0.3572,-0.1368),
    c(0.4576,-0.3157,0.4708,0.1410,-0.4479,-0.1685,0.3668,-0.2692,0.1332))

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 1.23, 
               highd_data = scaled_pbmc_data, 
               model_highd = scaled_pbmc_data_model, 
               trimesh_data = tr_from_to_df_pbmc, 
               axis_param = list(limits = 1,
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.48, 
                                 axis_pos_y = -0.48, 
                                 threshold = 0.082))

proj_obj1[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

@@

## Second projection
projection <- cbind(
    c(0.0734,0.6138,-0.4671,0.2589,0.1189,0.0778,-0.0819,0.0308,-0.5561),
    c(0.6625,0.0034,0.2412,0.3360,0.2463,0.1070,-0.2903,-0.4705,0.1293))

proj_obj2 <- get_projection(projection = projection, 
               proj_scale = 1.25, 
               highd_data = scaled_pbmc_data, 
               model_highd = scaled_pbmc_data_model, 
               trimesh_data = tr_from_to_df_pbmc, 
               axis_param = list(limits = 1.35, 
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.75, 
                                 axis_pos_y = -0.75, 
                                 threshold = 0.12))

proj_obj2[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

@@

<!--best choice-->

```{r}
#| label: tsne-pbmc-read-best
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_30.rds")
```

<!--Fit the best model and compute error-->
```{r}
#| label: tsne-num-bins-pbmc

## Compute hexbin parameters
num_bins_x_tsne_pbmc <- 22

algo_obj_pbmc <- fit_highd_model(
  highd_data = training_data_pbmc, 
  nldr_data = tsne_pbmc, 
  bin1 = num_bins_x_tsne_pbmc, 
  q = 0.1, 
  benchmark_highdens = 1)

tsne_pbmc_scaled_best <- algo_obj_pbmc$nldr_obj$scaled_nldr
tr_from_to_df_pbmc <- algo_obj_pbmc$trimesh_data
df_bin_centroids_pbmc <- algo_obj_pbmc$model_2d
df_bin_pbmc <- algo_obj_pbmc$model_highd

tsne_pbmc_scaled_best_with_cluster <- tsne_pbmc_scaled_best |>
  mutate(cluster = umap_pbmc_scaled_with_cluster$cluster)

trimesh_removed_pbmc_best <- ggplot() + 
  geom_point(
    data = tsne_pbmc_scaled_best_with_cluster,
    aes(
      x = tSNE1,
      y = tSNE2,
      colour = cluster
    ),
    alpha = 0.3#,
    #colour = "#999999"
  ) +
  geom_segment(data = tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("e1", cex = 2, c(0.08, 0.9)) +
  theme(
    aspect.ratio = 1
  )
```

```{r}
#| label: prep-pbmc-tsne-model-proj-best

df_b_pbmc <- df_bin_pbmc |>
  dplyr::filter(hexID %in% df_bin_centroids_pbmc$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_pbmc <- df_b_pbmc[match(df_bin_centroids_pbmc$hexID, df_b_pbmc$hexID),] |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data_pbmc <- bind_rows(data_pbmc, df_b_pbmc)
scaled_pbmc <- scale_data_manual(df_model_data_pbmc, "type") |>
  as_tibble()

scaled_pbmc_data <- scaled_pbmc |>
  filter(type == "data") |>
  select(-type)

scaled_pbmc_data_model <- scaled_pbmc |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-pbmc-tsne-best-proj
#| eval: false

df_model_data_pbmc_n <- bind_rows(df_b_pbmc, data_pbmc_n)

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = tr_from_to_df_pbmc$from,
                         lineTo = tr_from_to_df_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

## First projection
projection <- cbind(
  c(-0.4545,-0.5533,-0.1235,0.3984,-0.2065,-0.3586,-0.0088,0.3572,-0.1368),
  c(0.4576,-0.3157,0.4708,0.1410,-0.4479,-0.1685,0.3668,-0.2692,0.1332))

proj_obj1 <- get_projection(projection = projection, 
                            proj_scale = 1.23, 
                            highd_data = scaled_pbmc_data, 
                            model_highd = scaled_pbmc_data_model, 
                            trimesh_data = tr_from_to_df_pbmc, 
                            axis_param = list(limits = 1, 
                                              axis_scaled = 1,
                                              axis_pos_x = -0.48, 
                                              axis_pos_y = -0.48, 
                                              threshold = 0.082))

proj_obj1[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

@@

## Second projection
projection <- cbind(
  c(0.0734,0.6138,-0.4671,0.2589,0.1189,0.0778,-0.0819,0.0308,-0.5561),
  c(0.6625,0.0034,0.2412,0.3360,0.2463,0.1070,-0.2903,-0.4705,0.1293))

proj_obj2 <- get_projection(projection = projection, 
                            proj_scale = 1.25, 
                            highd_data = scaled_pbmc_data, 
                            model_highd = scaled_pbmc_data_model, 
                            trimesh_data = tr_from_to_df_pbmc, 
                            axis_param = list(limits = 1.35, 
                                              axis_scaled = 1,
                                              axis_pos_x = -0.75, 
                                              axis_pos_y = -0.75, 
                                              threshold = 0.12))

proj_obj2[["cluster"]] <- as.character(umap_pbmc_scaled_with_cluster$cluster)

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-rmse a) and the \\gD{} layout selected (@fig-pbmc-rmse e) by RMSE plot (@fig-pbmc-rmse) from the UMAP and tSNE with default (hyper-)~parameters. The PBMC3k data ($n =  2622$) has three close clusters in $9\\text{-}D$. Two \\gD{} projection from a tour on $9\\text{-}D$ of the model fit with @fig-pbmc-rmse a ($a_1 = 0.06$, $b = 616/86 (22, 28)$) shows three-well separated clusters with big separations. On the other hand, the model fit with @fig-pbmc-rmse e ($a_1 = 0.06$, $b = 704/227 (22, 32)$) shows three-well separated clusters with small separations. Therefore, @fig-pbmc-rmse e is more reasonable than @fig-pbmc-rmse a. Also, the fitted models in $9\\text{-}D$ help to see some unobserved patterns of the data: (i) dense points, and (ii) nonlinear clusters. Videos of the langevitour animations are available at <https://youtu.be/3VfK3M2gnZM> and <https://youtu.be/Es84bwQcndU> respectively."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

>>>>

@@

<!--Fit the best model for author suggestion and compute error-->
```{r}
#| label: hexbin-pbmc
#| 
umap_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/umap_pbmc_scaled_with_cluster.rds")

umap_tr_from_to_df_pbmc <- read_rds("data/pbmc3k/umap_tr_from_to_df_pbmc.rds")

trimesh_removed_pbmc <- ggplot() + 
  geom_point(
    data = umap_pbmc_scaled_with_cluster,
    aes(
      x = UMAP1,
      y = UMAP2,
      colour = cluster
    ),
    alpha = 0.2#,
    #size = 0.1#,
    #colour = '#a65628'
  )  +
  geom_segment(data = umap_tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("a1", c(0.08, 0.9), cex = 2) +
  theme(
    aspect.ratio = 1
  )
```

@@

## First projection
proj_obj1 <- read_rds("data/pbmc3k/pbmc_umap_proj_obj1.rds")

@@

## Second projection
proj_obj2 <- read_rds("data/pbmc3k/pbmc_umap_proj_obj2.rds")

@@

<!--best choice-->
<!--Fit the best model and compute error-->
```{r}
#| label: tsne-num-bins-pbmc
tsne_pbmc_scaled_best_with_cluster <- read_rds("data/pbmc3k/tsne_pbmc_scaled_best_with_cluster.rds")

tsne_tr_from_to_df_pbmc <- read_rds("data/pbmc3k/tsne_tr_from_to_df_pbmc.rds")

trimesh_removed_pbmc_best <- ggplot() + 
  geom_point(
    data = tsne_pbmc_scaled_best_with_cluster,
    aes(
      x = tSNE1,
      y = tSNE2,
      colour = cluster
    ),
    alpha = 0.3#,
    #colour = "#999999"
  ) +
  geom_segment(data = tsne_tr_from_to_df_pbmc, 
               aes(
                 x = x_from, 
                 y = y_from, 
                 xend = x_to, 
                 yend = y_to),
               colour = "#000000",
               linewidth = 1) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada")) +
  interior_annotation("e1", cex = 2, c(0.08, 0.9)) +
  theme(
    aspect.ratio = 1
  )
```

@@

## First projection
proj_obj1 <- read_rds("data/pbmc3k/pbmc_tsne_proj_obj1.rds")

@@

## Second projection
proj_obj2 <- read_rds("data/pbmc3k/pbmc_tsne_proj_obj2.rds")

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (@fig-pbmc-rmse a) and the \\gD{} layout selected (@fig-pbmc-rmse e) by RMSE plot (@fig-pbmc-rmse) from the UMAP and tSNE with default (hyper-)~parameters. The PBMC3k data ($n =  2622$) has three close clusters in $9\\text{-}D$. Two \\gD{} projection from a tour on $9\\text{-}D$ of the model fit with @fig-pbmc-rmse a ($a_1 = 0.06$, $b = 616/86 (22, 28)$) shows three-well separated clusters with big separations. On the other hand, the model fit with @fig-pbmc-rmse e ($a_1 = 0.06$, $b = 704/227 (22, 32)$) shows three-well separated clusters with small separations. Therefore, @fig-pbmc-rmse e is more reasonable than @fig-pbmc-rmse a. Also, the fitted models in $9\\text{-}D$ help to see some unobserved patterns of the data: (i) dense points, and (ii) nonlinear clusters. The links to the videos of langevitour animations are available in @tbl-links."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%
